A dynamic bioenergetic model for coral-\emph{Symbiodinium} symbioses and coral bleaching as an alternate stable state

### 2017-07-11 ms/ms.Rmd
revisions

In this dynamical system, both the coral host and algal symbiont acquire and use carbon and nitrogen to construct biomass. The symbiont fixes carbon through photosynthesis and receives nitrogen shared by the host, while the host acquires nitrogen from the environment and receives carbon shared by the symbiont. A graphical representation of the model is presented in Fig. 1, and each model flux and parameter is defined in Tables 1 and 2, respectively. We use C-moles (C-mol) as the unit of biomass for consistency with the rigorous mass balance of DEB theory: 1 C-mol is equivalent to the amount of biomass containing 1 mole of carbon atoms. Host biomass (expressed as C-mol H), symbiont biomass (expressed as C-mol S), and prey biomass (expressed as C-mol X) have fixed, but different, molar N:C ratios (Table 2). Biomass is produced from carbon and nitrogen by synthesizing units (SU), which are mathematical specifications of the formation of a product from two substrates; we use the "parallel complementary" formulation of @Kooijman:2010vd (page 105, Fig. 3.7) to specify these fluxes. The two state variables of this system are symbiont biomass and coral biomass; because resources are acquired proportionally to surface area, and surface area is proportional to volume, biomass increases exponentially during growth (indeed, corals grow exponentially [@Bak:1976bv]).

@@

Table 1. Model fluxes (mass-specific). Units are explained in the text.

@@

Table 2. Model parameters (units explained in the text). Justification and/or derivation of each parameter value along with supporting references are provided in the Supplementary Information.

@@

Coral biomass formation is then specified by a parallel complementary SU (formula in Fig. 3.7 of @Kooijman:2010vd]) that combines carbon and nitrogen, according to:

@@

The symbiont produces fixed carbon through photosynthesis, a process represented here by a single SU with two substrates: light (photons) and inorganic carbon (CO~2~). The amount of light absorbed by the symbiont depends on the scalar irradiance at the site of light absorption, which is modified substantially relative to external downwelling irradiance owing to multiple scattering by the coral skeleton and self-shading by surrounding symbionts [@Enriquez:2005p142; @Marcelino:2013hz]. We used data from @Marcelino:2013hz to empirically derive an amplification factor, $A$, indicating the ratio of internal scalar irradiance to external downwelling irradiance as a function of symbiont density (S:H biomass), which is specified as:

@@

[^2]: Note that recycling of symbiont biomass turnover ($rNS$ and $rCS$) only occurs based on the maintenance component of turnover (i.e., $j_{ST}^0$), and not the photodamage/bleaching component, as this loss represents biomass that is expelled from the host.
[^3]: See footnote 2.

@@

![Sensitivity analysis. Plots show the fractional change in steady state values of growth (solid lines) and S:H biomass (dashed lines) in response to fractional changes in default parameter values (see Table 2 for default values). Parameters are grouped by the processes in which they are involved. This sensitivity analysis was conducted at conditions typical for coral reef environments: low DIN (1e-7 M) and intermediate light (15 mol photons m^-2^ d^-2^), with prey density set to zero. Sensitivity analyses conducted other environmental conditions are presented in Figs S3-S7.](../img/Fig4.png)

@@

The prediction of higher growth when light is reduced indicates that growth is not limited by low light in winter, but is actually reduced by excess light in summer[^4], consistent with the experimental findings of @Schutter:2011ie. Seasonal summertime reductions in tissue biomass have also been well-documented in the field [@Fitt:2000p3792], along with reductions in net photosynthetic capacity [@MullerParker:1987cr]. Importantly, while light alone may drive seasonal dynamics in the ways discussed[^5], temperature fluctuations may attenuate or even reverse the effect of light as cooler winters depress metabolism; thus, the relative magnitude of fluctuation in temperature, light, and other factors may produce wide variability in the direction and magnitude of seasonal changes in growth and symbiont abundance, depending on location and microhabitat. Nevertheless, the seasonal variability predicted here (Fig. 5) is consistent with experimental and field observations for corals, and demonstrates the model's ability to predict dynamic behavior that mechanistically integrates multiple environmental drivers.

>>>>

In this dynamical system, both the coral host and algal symbiont acquire and use carbon and nitrogen to construct biomass. The symbiont fixes carbon through photosynthesis and receives nitrogen shared by the host, while the host acquires nitrogen from the environment and receives carbon shared by the symbiont. A graphical representation of the model is presented in Fig. 1, and each model flux and parameter is defined in Tables 1 and 2, respectively. We use C-moles (C-mol) as the unit of biomass for consistency with the rigorous mass balance of DEB theory: 1 C-mol is equivalent to the amount of biomass containing 1 mole of carbon atoms. Host biomass (expressed as C-mol H), symbiont biomass (expressed as C-mol S), and prey biomass (expressed as C-mol X) have fixed, but different, molar N:C ratios (Table 2), consistent with the assumption of strong homeostasis in DEB theory [@Kooijman:2010vd]. Biomass is produced from carbon and nitrogen by parallel complementary synthesizing units (SUs) -- mathematical specifications of the formation of a product by a metabolic network processing two complementary substrates in parallel (@Kooijman:2010vd, p. 105, Fig. 3.7). The two state variables in the model are symbiont biomass and coral biomass; because resources are acquired proportionally to surface area, and surface area is proportional to volume (corals are V1-morphs in a DEB context (@Kooijman:2010vd, p. 122)), biomass increases exponentially during growth (indeed, corals grow exponentially [@Bak:1976bv]).

@@

Table 1. Model fluxes (mass-specific). Units are explained in the text. (Abbreviations: CCMs=carbon concentrating mechanisms; NPQ=non-photochemical quenching; ROS=reactive oxygen species)

@@

Table 2. Model parameters (units explained in the text). Justification and/or derivation of each parameter value along with supporting references are provided in the Supplementary Information. (Abbreviations: CCMs=carbon concentrating mechanisms; NPQ=non-photochemical quenching; ROS=reactive oxygen species; DIN=dissolved inorganic nitrogen)

@@

Coral biomass formation is then specified by a parallel complementary SU (formula in Fig. 3.7 of @Kooijman:2010vd]). The general form of this equation is $(m^{-1} + x^{-1} + y^{-1} - (x + y)^{-1})^{-1}$, where $m$ is the maximum production flux, and $x$ and $y$ are the input fluxes of the two substrates. Coral biomass is thus produced from carbon and nitrogen according to:

@@

The symbiont produces fixed carbon through photosynthesis, a process represented here by a single SU with two substrates: light (photons) and inorganic carbon (CO~2~). The amount of light absorbed by the symbiont depends on the scalar irradiance at the site of light absorption, which is modified substantially relative to external downwelling irradiance owing to multiple scattering by the coral skeleton and self-shading by surrounding symbionts [@Enriquez:2005p142; @Marcelino:2013hz]. We used skeletal light amplification measurements from @Marcelino:2013hz to empirically derive an amplification factor, $A$ (Figure S1), indicating the ratio of internal scalar irradiance to external downwelling irradiance as a function of symbiont density (S:H biomass), which is specified as:

@@

[^2]: Note that recycling of symbiont biomass turnover ($r_{NS}$ and $r_{CS}$) only occurs based on the maintenance component of turnover (i.e., $j_{ST}^0$), and not the photodamage/bleaching component, as this loss represents biomass that is expelled from the host.
[^3]: See footnote 2.

@@

![Sensitivity analysis. Plots show the fractional change in steady state values of growth rate (solid lines) and S:H biomass (dashed lines) in response to fractional changes in default parameter values (see Table 2 for default values). Parameters are grouped by the processes in which they are involved. This sensitivity analysis was conducted at conditions typical for coral reef environments: low DIN (1e-7 M) and intermediate light (15 mol photons m^-2^ d^-2^), with prey density set to zero. Sensitivity analyses conducted other environmental conditions are presented in Figs S3-S7.](../img/Fig4.png)

@@

The prediction of higher growth when light is reduced indicates that growth is not limited by low light in winter, but is actually reduced by excess light in summer[^4], consistent with the experimental findings of @Schutter:2011ie. Seasonal summertime reductions in tissue biomass have also been well-documented in the field [@Fitt:2000p3792], along with reductions in net photosynthetic capacity [@MullerParker:1987cr]. Importantly, while light alone may drive seasonal dynamics in the ways discussed[^5], temperature fluctuations may attenuate or even reverse the effect of light as cooler winters depress metabolism. Seasonal changes in nutrients associated with fertilization use and runoff during growing seasons may also impact these dynamics. Thus, the relative magnitude of fluctuation in temperature, light, and nutrients may produce wide variability in the direction and magnitude of seasonal changes in growth and symbiont abundance, depending on location and microhabitat. Nevertheless, the seasonal variability predicted here in response to light (Fig. 5) is consistent with experimental and field observations for corals, and demonstrates the model's ability to predict dynamic behavior that mechanistically integrates multiple environmental drivers.

### 2017-07-11 ms/ms.tex
revisions

In this dynamical system, both the coral host and algal symbiont acquire
and use carbon and nitrogen to construct biomass. The symbiont fixes
carbon through photosynthesis and receives nitrogen shared by the host,
while the host acquires nitrogen from the environment and receives
carbon shared by the symbiont. A graphical representation of the model
is presented in Fig. 1, and each model flux and parameter is defined in
Tables 1 and 2, respectively. We use C-moles (C-mol) as the unit of
biomass for consistency with the rigorous mass balance of DEB theory: 1
C-mol is equivalent to the amount of biomass containing 1 mole of carbon
atoms. Host biomass (expressed as C-mol H), symbiont biomass (expressed
as C-mol S), and prey biomass (expressed as C-mol X) have fixed, but
different, molar N:C ratios (Table 2). Biomass is produced from carbon
and nitrogen by synthesizing units (SU), which are mathematical
specifications of the formation of a product from two substrates; we use
the ``parallel complementary'' formulation of Kooijman (2010) (page 105,
Fig. 3.7) to specify these fluxes. The two state variables of this
system are symbiont biomass and coral biomass; because resources are
acquired proportionally to surface area, and surface area is
proportional to volume, biomass increases exponentially during growth
(indeed, corals grow exponentially (Bak 1976)).

@@

Coral biomass formation is then specified by a parallel complementary SU
(formula in Fig. 3.7 of Kooijman (2010){]}) that combines carbon and
nitrogen, according to:

@@

The symbiont produces fixed carbon through photosynthesis, a process
represented here by a single SU with two substrates: light (photons) and
inorganic carbon (CO\textsubscript{2}). The amount of light absorbed by
the symbiont depends on the scalar irradiance at the site of light
absorption, which is modified substantially relative to external
downwelling irradiance owing to multiple scattering by the coral
skeleton and self-shading by surrounding symbionts (Enríquez, Méndez,
and Iglesias-Prieto 2005; Marcelino et al. 2013). We used data from
Marcelino et al. (2013) to empirically derive an amplification factor,
\(A\), indicating the ratio of internal scalar irradiance to external
downwelling irradiance as a function of symbiont density (S:H biomass),
which is specified as:

@@

along with a fixed proportion of CO\textsubscript{2} generated by
symbiont biomass turnover\footnote{Note that recycling of symbiont
  biomass turnover (\(rNS\) and \(rCS\)) only occurs based on the
  maintenance component of turnover (i.e., \(j_{ST}^0\)), and not the
  photodamage/bleaching component, as this loss represents biomass that
  is expelled from the host.} and formation:

@@

\begin{figure}[htbp]
\centering
\includegraphics{../img/Fig4.png}
\caption{Sensitivity analysis. Plots show the fractional change in
steady state values of growth (solid lines) and S:H biomass (dashed
lines) in response to fractional changes in default parameter values
(see Table 2 for default values). Parameters are grouped by the
processes in which they are involved. This sensitivity analysis was
conducted at conditions typical for coral reef environments: low DIN
(1e-7 M) and intermediate light (15 mol photons m\textsuperscript{-2}
d\textsuperscript{-2}), with prey density set to zero. Sensitivity
analyses conducted other environmental conditions are presented in Figs
S3-S7.}
\end{figure}

@@

The prediction of higher growth when light is reduced indicates that
growth is not limited by low light in winter, but is actually reduced by
excess light in summer\footnote{at the light levels indicated in Stimson
  (1997). Note that if light levels were reduced throughout the year
  (e.g.~for a coral at greater depth) such that light did not cause
  photo-oxidative stress in summer but became limiting to photosynthesis
  in winter, the S:H ratio would still increase in winter, but growth
  would decrease; the latter scenario is predicted both by the present
  model as well as that of Muller et al. (2009), since photo-oxidative
  stress does not become relevant.}, consistent with the experimental
findings of Schutter et al. (2011). Seasonal summertime reductions in
tissue biomass have also been well-documented in the field (Fitt et al.
2000), along with reductions in net photosynthetic capacity
(Muller-Parker 1987). Importantly, while light alone may drive seasonal
dynamics in the ways discussed\footnote{See footnote 4.}, temperature
fluctuations may attenuate or even reverse the effect of light as cooler
winters depress metabolism; thus, the relative magnitude of fluctuation
in temperature, light, and other factors may produce wide variability in
the direction and magnitude of seasonal changes in growth and symbiont
abundance, depending on location and microhabitat. Nevertheless, the
seasonal variability predicted here (Fig. 5) is consistent with
experimental and field observations for corals, and demonstrates the
model's ability to predict dynamic behavior that mechanistically
integrates multiple environmental drivers.

>>>>

In this dynamical system, both the coral host and algal symbiont acquire
and use carbon and nitrogen to construct biomass. The symbiont fixes
carbon through photosynthesis and receives nitrogen shared by the host,
while the host acquires nitrogen from the environment and receives
carbon shared by the symbiont. A graphical representation of the model
is presented in Fig. 1, and each model flux and parameter is defined in
Tables 1 and 2, respectively. We use C-moles (C-mol) as the unit of
biomass for consistency with the rigorous mass balance of DEB theory: 1
C-mol is equivalent to the amount of biomass containing 1 mole of carbon
atoms. Host biomass (expressed as C-mol H), symbiont biomass (expressed
as C-mol S), and prey biomass (expressed as C-mol X) have fixed, but
different, molar N:C ratios (Table 2), consistent with the assumption of
strong homeostasis in DEB theory (Kooijman 2010). Biomass is produced
from carbon and nitrogen by parallel complementary synthesizing units
(SUs) -- mathematical specifications of the formation of a product by a
metabolic network processing two complementary substrates in parallel
(Kooijman (2010), p.~105, Fig. 3.7). The two state variables in the
model are symbiont biomass and coral biomass; because resources are
acquired proportionally to surface area, and surface area is
proportional to volume (corals are V1-morphs in a DEB context (Kooijman
(2010), p.~122)), biomass increases exponentially during growth (indeed,
corals grow exponentially (Bak 1976)).

@@

Coral biomass formation is then specified by a parallel complementary SU
(formula in Fig. 3.7 of Kooijman (2010){]}). The general form of this
equation is \((m^{-1} + x^{-1} + y^{-1} - (x + y)^{-1})^{-1}\), where
\(m\) is the maximum production flux, and \(x\) and \(y\) are the input
fluxes of the two substrates. Coral biomass is thus produced from carbon
and nitrogen according to:

@@

The symbiont produces fixed carbon through photosynthesis, a process
represented here by a single SU with two substrates: light (photons) and
inorganic carbon (CO\textsubscript{2}). The amount of light absorbed by
the symbiont depends on the scalar irradiance at the site of light
absorption, which is modified substantially relative to external
downwelling irradiance owing to multiple scattering by the coral
skeleton and self-shading by surrounding symbionts (Enríquez, Méndez,
and Iglesias-Prieto 2005; Marcelino et al. 2013). We used skeletal light
amplification measurements from Marcelino et al. (2013) to empirically
derive an amplification factor, \(A\) (Figure S1), indicating the ratio
of internal scalar irradiance to external downwelling irradiance as a
function of symbiont density (S:H biomass), which is specified as:

@@

along with a fixed proportion of CO\textsubscript{2} generated by
symbiont biomass turnover\footnote{Note that recycling of symbiont
  biomass turnover (\(r_{NS}\) and \(r_{CS}\)) only occurs based on the
  maintenance component of turnover (i.e., \(j_{ST}^0\)), and not the
  photodamage/bleaching component, as this loss represents biomass that
  is expelled from the host.} and formation:

@@

\begin{figure}[htbp]
\centering
\includegraphics{../img/Fig4.png}
\caption{Sensitivity analysis. Plots show the fractional change in
steady state values of growth rate (solid lines) and S:H biomass (dashed
lines) in response to fractional changes in default parameter values
(see Table 2 for default values). Parameters are grouped by the
processes in which they are involved. This sensitivity analysis was
conducted at conditions typical for coral reef environments: low DIN
(1e-7 M) and intermediate light (15 mol photons m\textsuperscript{-2}
d\textsuperscript{-2}), with prey density set to zero. Sensitivity
analyses conducted other environmental conditions are presented in Figs
S3-S7.}
\end{figure}

@@

The prediction of higher growth when light is reduced indicates that
growth is not limited by low light in winter, but is actually reduced by
excess light in summer\footnote{at the light levels indicated in Stimson
  (1997). Note that if light levels were reduced throughout the year
  (e.g.~for a coral at greater depth) such that light did not cause
  photo-oxidative stress in summer but became limiting to photosynthesis
  in winter, the S:H ratio would still increase in winter, but growth
  would decrease; the latter scenario is predicted both by the present
  model as well as that of Muller et al. (2009), since photo-oxidative
  stress does not become relevant.}, consistent with the experimental
findings of Schutter et al. (2011). Seasonal summertime reductions in
tissue biomass have also been well-documented in the field (Fitt et al.
2000), along with reductions in net photosynthetic capacity
(Muller-Parker 1987). Importantly, while light alone may drive seasonal
dynamics in the ways discussed\footnote{See footnote 4.}, temperature
fluctuations may attenuate or even reverse the effect of light as cooler
winters depress metabolism. Seasonal changes in nutrients associated
with fertilization use and runoff during growing seasons may also impact
these dynamics. Thus, the relative magnitude of fluctuation in
temperature, light, and nutrients may produce wide variability in the
direction and magnitude of seasonal changes in growth and symbiont
abundance, depending on location and microhabitat. Nevertheless, the
seasonal variability predicted here in response to light (Fig. 5) is
consistent with experimental and field observations for corals, and
demonstrates the model's ability to predict dynamic behavior that
mechanistically integrates multiple environmental drivers.
